A test of diversifying selection for a trait from within and between species variations

### 2023-06-06 manuscript/main.tex
Updated manuscript

Determining whether a particular trait is neutrally evolving or under a particular regime of selection has been a long-standing goal in evolutionary biology.
Fundamentally, distinguishing neutral evolution from selection requires to determine which mode of selection is supported from the observed variation of traits and sequences.
This variation of traits can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
All these scales requires different assumptions and methodology, and the endeavor to determine the mode of selection for a given trait has thus incorporated theories, methods and developments across different fields of evolutionary biology such as quantitative-genetics, population-genetics, phylogenetics and comparative genomics\cite{lynch_genetics_1998, walsh_evolution_2018}.

@@

For a given trait, its genetic architecture is defined by the number of loci encoding the trait ($\NbrLoci$) and the random effect of a mutation on the trait ($a$).
New mutations are thus generating variance for such a trait.
At the individual level, the mutational variance ($\VarMutation$) is the rate at which new mutations contributes to the trait variance per generation.
As shown in \textcite{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of the mutation rate per loci per generation ($\MutationRate$), as:
\begin{gather}
    \VarMutation = 2 \MutationRate \Multiply \NbrLoci \Multiply \E \left[ a^2 \right] \label{eq:var-mutation}.
\end{gather}

At the population level, mutations supply new variants in the population, while genetic random drift depletes standing variation.
As shown in \textcite{lynch_mutation_1998}, if the trait is neutral, at equilibrium between mutation and drift, the genetic variance in a population ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individual in the population ($\Ne$) as:
\begin{align}
    \VarGenetic & =  2 \Ne \Multiply \VarMutation, \\
    & = 4 \Ne \Multiply \MutationRate \Multiply \NbrLoci \Multiply \E \left[ a^2 \right] \text{ from eq.~\ref{eq:var-mutation}}\label{eq:var-genetic}.
\end{align}

Moreover, for any neutral genomic region of interest, the genetic diversity, called $\pi$, is measured as the number of mutation segregating in the population divided by the size of the region.
Any of the segregating mutation will eventually reach fixation or extinction due to genetic drift, such that the amount of segregating mutations is also a balance between mutations and drift.
As shown in \textcite{tajima_statistical_1989}, $\pi$ is thus a function of the mutation rate per loci per generation ($\MutationRate$) and the effective population size ($\Ne$), given as:
\begin{gather}
    \pi = 4 \Ne \Multiply \MutationRate \label{eq:genetic-diversity}.
\end{gather}

Altogether, we define define $\RatePop$ as the ratio of genetic variance of the trait ($\VarGenetic$) over genetic diversity of any neutral genomic region of interest ($\pi$).
This ratio allows to the discard the effect of $\Ne$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RatePop$ as:
\begin{align}
    \RatePop & \defEqual \frac{\VarGenetic }{\pi}, \\
    & = \frac{4 \Ne \Multiply \MutationRate \Multiply \NbrLoci \Multiply \E \left[ a^2 \right]}{4 \Ne \Multiply \MutationRate} \text{ from eq.~\ref{eq:var-mutation} and~\ref{eq:genetic-diversity}}, \\
    & = \NbrLoci \Multiply \E \left[ a^2 \right]. \label{eq:rate-pop}
\end{align}

Additionally, the genetic variance is equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by heredity ($h^2$), giving also $\RatePop$ as:
\begin{align}
    \RatePop & = \frac{h^2 \Multiply \VarPhenotype }{\pi }. \label{eq:rate-pheno-pop}
\end{align}

@@

Along a lineage, we denote $\Trait$ as the mean value of the trait over the individuals in the species.
If the trait is neutral and encoded by many loci, $\Trait$ evolves as Brownian process\cite{hansen_translating_1996}.
After the species has evolved for $\Time$ generations, the variance of $\Trait$, called $\VarPhy$, is given by \textcite{hansen_translating_1996} as:
\begin{align}
    \VarPhy & = \frac{\VarGenetic}{\Ne} \Multiply \Time \\
    & = 4 \Time \Multiply \MutationRate \Multiply \NbrLoci \Multiply \E \left[ a^2 \right] \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}},
\end{align}

Moreover, for any neutral genomic region of interest, out of all neutral mutations that originated, some will we eventually reach fixation due to genetic drift, resulting in a substitution at the species level.
For a neutral mutation, its probability of fixation ($\pfix$) is $1/2\Ne$, as shown in \textcite{kimura_probability_1962}.
As reviewed in \textcite{mccandlish_modeling_2014}, we can thus derive the substitution rate per generation, denoted $\SubRate$ as the number of mutations per generation ($2\Ne \Multiply \MutationRate$) multiplied by the probability of fixation for each of these newly arisen mutations $\pfix$, giving:
\begin{align}
    \SubRate & = 2 \Ne \Multiply \MutationRate \Multiply \pfix, \\
    & = 2 \Ne  \Multiply \MutationRate  \Multiply \dfrac{1}{2\Ne}, \\
    & = \MutationRate. \label{eq:substitution-rate}
\end{align}
That is, if mutations are neutral, the rate with which a new neutral allele are fixed in a genomic region of interest equals the rate with which new mutations arise per generation for the same genomic region\cite{kimura_evolutionary_1968}.

Thus, after having evolved for $\Time$ generations, the fraction of the genomic region of interest that generated a substitution, called $d$, will be given as the number of generations ($\Time$) multiplied by the substitution rate per generation ($\SubRate$), given as:
\begin{align}
    d & = \Time \Multiply \SubRate \\
    & = \Time \Multiply \MutationRate \text{ from eq.~\ref{eq:substitution-rate}}. \label{eq:distance}
\end{align}
Altogether, we defined $\RatePhy$ as the variance in the mean trait value ($\VarPhy$) normalized by the genetic distance of any neutral genomic region of interest ($d$).
This ratio allows to the discard the effect of $\Time$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RatePhy$ as:
\begin{align}
    \RatePhy & \defEqual \frac{\VarPhy}{4 d},\\
    & = \frac{4 \Time \Multiply \MutationRate \Multiply \NbrLoci \Multiply \E \left[ a^2 \right]}{4 \Time \Multiply \MutationRate}\text{ from eq.~\ref{eq:covar} and~\ref{eq:distance}}, \\
    & = \NbrLoci \Multiply \E \left[ a^2 \right] \label{eq:rate-phy}.
\end{align}

@@

 The variability between individuals or between species can be obtained for both quantitative trait and genomic sequences.
 At the population level, the variability of the trait between individuals allows to compute $\VarGenetic$, and the variability and of any genomic region neutrally evolving allow to compute $\pi$.
Combined as a ratio, we obtain $\RatePop$ at the population level, which depends solely on the genetic architecture of the trait (as $\NbrLoci \Multiply \E \left[ a^2 \right]$).
At the phylogenetic level, the variability of the mean trait value between species allows to compute $\VarPhy$, and the variability and of any genomic region neutrally evolving allows to compute $d$.
Similarly, combined as a ratio, we obtain $\RatePhy$ at the phylogenetic scale, which also depends solely on the genetic architecture of the trait.
Altogether, we thus have:
\begin{gather}
    \RatePop = \RatePhy \text{ from eq.~\ref{eq:rate-pop} and~\ref{eq:rate-phy}} , \\
    \Rightarrow \frac{\RatePhy}{\RatePop} = \frac{\VarPhy \Multiply \pi}{ \VarGenetic \Multiply 4 d }  = 1.
\end{gather}
Altogether, $\RatePhy / \RatePop$ is our neutrality index, which should identify to $1$ for a trait neutrally evolving.
Importantly, $\RatePhy$ and $\RatePop$ can both be estimated using quantitative trait and genomic sequences within and between species, while neither the mutation rate $\MutationRate$ nor the effective population size $\Ne$ need to be estimated.

@@

\subsubsection{Maximum likelihood estimate}
For each species of interest with inter-individual data available, $\RatePop$ can be estimated as the ratio of genetic variance of a trait over genetic diversity of neutrally evolving sequences.
Averaged across all species we obtain the mean as:
\begin{gather}
    \EstRatePop = \left< \frac{\VarGenetic}{\pi} \right> = \left< \frac{h^2 \Multiply \VarPhenotype }{\pi } \right> \label{eq:estimated-rate-pop}
\end{gather}

At the phylogenetic scale, for $\NbrSpecies$ taxa in the tree, $\Cr$ is the $\NbrSpecies \times \NbrSpecies$ contrast matrix computed from the branch lengths ($d$) and the topology of the phylogenetic tree.
The contrast matrix is symmetric, along the diagonal are the total distances ($d_{\Spi}$) from the root of the tree to each taxon ($\Spi$).
The off-diagonal elements $C_{\Spi,\Spj}$ are the total branch lengths shared by particular pairs of taxa $\Spi$ and $\Spj$, computed from the distance between the pair ($d_{\Spi,\Spj}$) as:
\begin{gather}
    C_{\Spi,\Spj} = d_{\mathcal{A}(\Spi, \Spj)} = \frac{d_{\Spi} + d_{\Spj} - d_{\Spi,\Spj}}{2}
\end{gather}
Then, the root state for the character, called $\RootTrait$, can be estimated from the $\NbrSpecies \times 1$ vector of mean trait values for tip species in the tree, called $\VecTrait$, as a maximum likelihood estimate\cite{omeara_testing_2006}, given as:
\begin{gather}
    \RootTrait = \left( \VecOne\tr \Cr\inv \VecOne \right)\inv \left( \VecOne\tr \Cr\inv \VecTrait \right), \label{eq:estimated-root-trait}
\end{gather}
where $\VecOne$ is an $\NbrSpecies \times 1$ column vector of ones.

Finally, the normalized mutational variance at the phylogenetic scale ($\EstRatePhy$) is estimated\cite{omeara_testing_2006} as:
\begin{gather}
    \EstRatePhy = \frac{1}{4}\frac{\left( \VecTrait -  \RootTrait \VecOne \right)\tr \Cr\inv \left( \VecTrait -  \RootTrait  \VecOne  \right)}{\NbrSpecies - 1}. \label{eq:estimated-rate-phy}
\end{gather}
Importantly, the sequence from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow to normalize for diversity driven by mutation rate and population size.

@@

We will assume that the trait genotypic value is encoded by $\NbrLoci$ loci, with each locus contributing ly to the genotypic value, these effect are Gamma distributed with mean $a$ (Fig.~\ref{fig:simulator}).
We will follow a Wright-Fisher model with mutation, selection and drift to formalise the evolutionary dynamics of the system.
Assuming a panmictic population of size $\Ne$ and with non-overlapping generations, we can simulate explicitly each generation along a phylogenetic tree.
Mutations will be based on a user-defined nucleotide matrix, while drift will be modelled by the multinomial resampling of the currently segregating alleles.
The changes in the mutation rate per generation, the generation time and $\Ne$ along the lineages will be modelled by a geometric Brownian multivariate and an Ornstein-Uhlenbeck process to test different processes of evolution of these parameters.
The instant value of log-$\Ne$ can also be modelled as a sum of a geometric Brownian process and an Ornstein-Uhlenbeck process.
The geometric Brownian motion accounts for long-term fluctuations, while the Ornstein-Uhlenbeck introduces short-term fluctuations.
The simulations will start from an initial sequence at equilibrium at the root of the tree and, at each node, the process will be split until it finally reaches the leaves of the tree.
We will introduce selection as a one dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and optimal phenotype \cite{tenaillon_utility_2014,blanquart_epistasis_2016}.
The exact functional phenotype-fitness map depends on $2$ external parameters controlling for strength of selection and optimum phenotype.
Mutations are seen has displacement of the phenotype in the multidimensional space.
Beneficial mutations will move the phenotype closer to the optimum, while deleterious mutations will move it further away.
Stabilizing selection will be implemented by means of a fixed optimum phenotype, while diversifying selection will be implemented by means of a moving optimum phenotype, changing along the phylogeny.

@@

\begin{table*}[t!]
    \centering
    \begin{adjustbox}{width = 0.5\textwidth}
    \begin{tabular}{|c||c|c|}
            \hline
            & Coding sequences & Quantitative traits \\ \hline \hline
            Adaptation / Diversifying selection & $\dnds > \pnps$ & $ \RatePhy > \RatePop  $ \\ \hline
            Neutral regime of evolution & $\dnds = \pnps$ & $ \RatePhy = \RatePop  $ \\ \hline
            Purifying / Stabilizing selection & $\dnds < \pnps$ & $ \RatePhy < \RatePop  $ \\ \hline
    \end{tabular}
    \end{adjustbox}
    \caption{
        The relationship between trait variation between and within species can be used as method to detect whether traits are evolving under a neutral model of evolution.
        Similarly, in molecular evolution, non-synonymous divergence ($\dn$) and polymorphism $\pn$ are normalized by synonymous divergence ($\ds$) and polymorphism ($\ps$) to produce a neutrality index\cite{mcdonald_adaptative_1991, fay_evaluating_2008}.
    }
    \label{table:unfolded-MK}
\end{table*}

Importantly, the genetic sequences from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow to normalize for diversity driven by mutation rate and population size.
Analogously to the \textcite{mcdonald_adaptative_1991} test devised for protein-coding DNA sequences, a category of neutral mutations (synonymous) is used to determine the neutral expectation, and the inflation of divergence is compared to polymorphism within species.
Altogether, our neutrality index for a quantitative trait leverages the data for any number of species, and takes advantage of the signal over the whole phylogeny, while at the same time taking into account phylogenetic inertia and addressing the non-independence between species (see fig.~\ref{fig:methods}).
This statistic can be obtained as a maximum likelihood estimate, from eq.~\ref{eq:estimated-rate-pop} and~\ref{eq:estimated-rate-phy}.

@@

In this study, we proposed a neutrality index for a quantitative trait.
At the phylogenetic scale, trait variation between species is normalized by sequence divergence obtained from a neutral set of markers.
Similarly, trait variation within species is normalized by sequence polymorphism obtained from a neutral set of markers.
Our neutrality index for a trait is the ratio of between species variation to within species variation.
Such neutrality index can be compared to 1 to seek a statistical deviation from a neutral trait.
In other words, our statistical test does to seeks to find traits with extremal values in comparison to a pool of traits.
Instead, our procedure can be applied to a single trait, estimating the neutrality index and giving a statistical test for departure from the null model of neutral evolution.
Technically, the neutrality index is estimated either as a maximum likelihood point estimate, or as mean posterior from a Bayesian implementation also giving access to the posterior credible interval allowing to test for a departure from a neutrally evolving trait.
Importantly, our neutrality index leverages sequence divergence and polymorphism, as well as the heritability of traits.
We argue that the main novelty of this study is being able to use sequence divergence and polymorphism to normalize the variation of trait between and within species.
By doing so, our test is not sensitive to the assumption that population sizes ($\Ne$) and mutation rates ($\MutationRate$) are constant across the phylogeny, not requiring to estimate divergence time\cite{litsios_effects_2012}.
Instead, the normalization by sequence divergence and polymorphism automatically absorbs these changes\cite{seo_estimating_2004}.
Contrarily, we argue that the main drawback of our method is that the heritability ($h^2$) of a trait is required if one has access only to the phenotypic variance ($\VarPhenotype$) and not the genetic variance ($\VarGenetic$), which is empirically hard to measure for any trait.

% What about the other methods? What are we testing for?
At the phylogenetic scale, the change in mean trait value across different species is typically used to test for neutrality for a quantitative trait.
These models are seeking deviation from a Brownian process\cite{catalan_drift_2019}, or from an Ornstein–Uhlenbeck process accounting for within-species variation\cite{rohlfs_phylogenetic_2015}.
For example, the analysis on the transcriptome across different individuals and species concluded that the expression level of 81\% of genes are under a neutral regime and 16\% under directional selection\cite{catalan_drift_2019}.
However, the alternative model of a varying directional or stabilizing selection was not implemented, and thus can not be ruled out by this analysis.
With our statistical test, we can set a threshold for neutrality and thus are able to make the difference between a trait evolving neutrally as a Brownian process or a trait under selection where the optimal is moving as a Brownian.
Our diversity index can thus allow to revisit these studies, assuming we have access to genomic datasets to estimate divergence ($d$) and polymorphism ($\pi$), as well as estimate for heritabilities ($h^2$).

@@

>>>>

Determining whether a particular trait is neutrally evolving or under a particular regime of selection has been a long-standing goal in evolutionary biology.
Fundamentally, distinguishing neutral evolution from selection requires determining which mode of selection is supported by the observed variation of traits and sequences.
Trait variation can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
All these scales require different assumptions and methodologies, and the endeavor to determine the mode of selection for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative-genetics, population-genetics, phylogenetics and comparative genomics\cite{lynch_genetics_1998, walsh_evolution_2018}.

@@

For a given trait, its genetic architecture is defined by the number of loci encoding the trait ($\NbrLoci$) and the random effect of a mutation on the trait ($a$).
New mutations are thus generating variance for such a trait.
At the individual level, the mutational variance ($\VarMutation$) is the rate at which new mutations contribute to the trait variance per generation.
As shown in \textcite{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of the mutation rate per loci per generation ($\MutationRate$), as:
\begin{gather}
    \VarMutation = 2 \MutationRate \Multiply \RateMut \label{eq:var-mutation}.
\end{gather}

At the population level, mutations supply new variants in the population, while genetic random drift depletes standing variation.
As shown in \textcite{lynch_mutation_1998}, if the trait is neutral, at equilibrium between mutation and drift, the genetic variance in a population ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individuals in the population ($\Ne$) as:
\begin{align}
    \VarGenetic & =  2 \Ne \Multiply \VarMutation, \\
    & = 4 \Ne \Multiply \MutationRate \Multiply \RateMut \text{ from eq.~\ref{eq:var-mutation}}\label{eq:var-genetic}.
\end{align}

Moreover, for any neutral genomic region of interest, the genetic diversity, called $\pi$, is measured as the number of mutation segregating in the population divided by the size of the region.
Any of the segregating mutations will eventually reach fixation or extinction due to genetic drift, such that the amount of segregating mutations is also a balance between mutations and drift.
As shown in \textcite{tajima_statistical_1989}, $\pi$ is thus a function of the mutation rate per loci per generation ($\MutationRate$) and the effective population size ($\Ne$), given as:
\begin{gather}
    \pi = 4 \Ne \Multiply \MutationRate \label{eq:genetic-diversity}.
\end{gather}

Altogether, we define $\RatePop$ as the ratio of genetic variance of the trait ($\VarGenetic$) over genetic diversity of any neutral genomic region of interest ($\pi$).
This ratio allows discarding the effect of $\Ne$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RatePop$ as:
\begin{align}
    \RatePop & \defEqual \frac{\VarGenetic }{\pi}, \\
    & = \frac{4 \Ne \Multiply \MutationRate \Multiply \RateMut}{4 \Ne \Multiply \MutationRate} \text{ from eq.~\ref{eq:var-mutation} and~\ref{eq:genetic-diversity}}, \\
    & = \RateMut. \label{eq:rate-pop}
\end{align}

Additionally, the genetic variance is equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by heredity ($\Heredity$), giving also $\RatePop$ as:
\begin{align}
    \RatePop & = \frac{\Heredity \Multiply \VarPhenotype }{\pi }. \label{eq:rate-pheno-pop}
\end{align}

@@

Along a lineage, we denote $\MeanTrait$ as the mean value of the trait over the individuals in the species.
If the trait is neutral and encoded by many loci, $\MeanTrait$ evolves as Brownian process\cite{hansen_translating_1996}.
After the species has evolved for $\Time$ generations, the variance of $\MeanTrait$, called $\VarPhy$, is given by \textcite{hansen_translating_1996} as:
\begin{align}
    \VarPhy & = \frac{\VarGenetic}{\Ne} \Multiply \Time \\
    & = 4 \Time \Multiply \MutationRate \Multiply \RateMut \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}},
\end{align}

Moreover, for any neutral genomic region of interest, out of all neutral mutations that originated, some will eventually reach fixation due to genetic drift, resulting in a substitution at the species level.
For a neutral mutation, its probability of fixation ($\pfix$) is $1/2\Ne$, as shown in \textcite{kimura_probability_1962}.
As reviewed in \textcite{mccandlish_modeling_2014}, we can thus derive the substitution rate per generation, denoted $\SubRate$ as the number of mutations per generation ($2\Ne \Multiply \MutationRate$) multiplied by the probability of fixation for each of these newly arisen mutations $\pfix$, giving:
\begin{align}
    \SubRate & = 2 \Ne \Multiply \MutationRate \Multiply \pfix, \\
    & = 2 \Ne  \Multiply \MutationRate  \Multiply \dfrac{1}{2\Ne}, \\
    & = \MutationRate. \label{eq:substitution-rate}
\end{align}
That is, if mutations are neutral, the rate with which a new neutral allele are fixed in a genomic region of interest equals the rate with which new mutations arise per generation for the same genomic region\cite{kimura_evolutionary_1968}.

Thus, after having evolved for $\Time$ generations, the fraction of the genomic region of interest that generated a substitution, called $d$, will be given as the number of generations ($\Time$) multiplied by the substitution rate per generation ($\SubRate$), given as:
\begin{align}
    d & = \Time \Multiply \SubRate \\
    & = \Time \Multiply \MutationRate \text{ from eq.~\ref{eq:substitution-rate}}. \label{eq:distance}
\end{align}
Altogether, we defined $\RatePhy$ as the variance in the mean trait value ($\VarPhy$) normalized by the genetic distance of any neutral genomic region of interest ($d$).
This ratio allows to the discard the effect of $\Time$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RatePhy$ as:
\begin{align}
    \RatePhy & \defEqual \frac{\VarPhy}{4 d},\\
    & = \frac{4 \Time \Multiply \MutationRate \Multiply \RateMut}{4 \Time \Multiply \MutationRate}\text{ from eq.~\ref{eq:covar} and~\ref{eq:distance}}, \\
    & = \RateMut \label{eq:rate-phy}.
\end{align}

@@

 The variability between individuals or between species can be obtained for both quantitative trait and genomic sequences.
 At the population level, the variability of the trait between individuals allows to compute $\VarGenetic$, and the variability and of any genomic region neutrally evolving allow to compute $\pi$.
Combined as a ratio, we obtain $\RatePop$ at the population level, which depends solely on the genetic architecture of the trait (as $\RateMut$).
At the phylogenetic level, the variability of the mean trait value between species allows to compute $\VarPhy$, and the variability and of any genomic region neutrally evolving allows to compute $d$.
Similarly, combined as a ratio, we obtain $\RatePhy$ at the phylogenetic scale, which also depends solely on the genetic architecture of the trait.
Altogether, we thus have:
\begin{gather}
    \RatePop = \RatePhy \text{ from eq.~\ref{eq:rate-pop} and~\ref{eq:rate-phy}} , \\
    \Rightarrow \frac{\RatePhy}{\RatePop} = \frac{\VarPhy \Multiply \pi}{ \VarGenetic \Multiply 4 d }  = 1.
\end{gather}
Altogether, $\RatePhy / \RatePop$ is our neutrality index, which should identify to $1$ for a trait neutrally evolving.
Importantly, $\RatePhy$ and $\RatePop$ can both be estimated using quantitative trait and genomic sequences within and between species, while neither the mutation rate $\MutationRate$ nor the effective population size $\Ne$ need to be estimated.

@@

\subsubsection{Maximum likelihood estimate}
At the phylogenetic scale, for $\NbrSpecies$ taxa in the tree, $\Cr$ is the $\NbrSpecies \times \NbrSpecies$ contrast matrix computed from the branch lengths ($d$) and the topology of the phylogenetic tree.
The contrast matrix is symmetric, along the diagonal are the total distances ($d_{\Spi}$) from the root of the tree to each taxon ($\Spi$).
The off-diagonal elements $C_{\Spi,\Spj}$ are the total branch lengths shared by particular pairs of taxa $\Spi$ and $\Spj$, computed from the distance between the pair ($d_{\Spi,\Spj}$) as:
\begin{gather}
    C_{\Spi,\Spj} = d_{\mathcal{A}(\Spi, \Spj)} = \frac{d_{\Spi} + d_{\Spj} - d_{\Spi,\Spj}}{2}
\end{gather}
Then, the root state for the character, called $\RootTrait$, can be estimated from the $\NbrSpecies \times 1$ vector of mean trait values for tip species in the tree, called $\VecTrait$, as a maximum likelihood estimate\cite{omeara_testing_2006}, given as:
\begin{gather}
    \RootTrait = \left( \VecOne\tr \MultiplyMatrix \Cr\inv \MultiplyMatrix \VecOne \right)\inv \Multiply \left( \VecOne\tr \MultiplyMatrix \Cr\inv \MultiplyMatrix \VecTrait \right), \label{eq:estimated-root-trait}
\end{gather}
where $\VecOne$ is an $\NbrSpecies \times 1$ column vector of ones.

Finally, the normalized mutational variance at the phylogenetic scale ($\EstRatePhy$) is estimated\cite{omeara_testing_2006} as:
\begin{gather}
    \EstRatePhy = \frac{1}{4}\frac{\left( \VecTrait -  \RootTrait \Multiply \VecOne \right)\tr \MultiplyMatrix \Cr\inv \MultiplyMatrix \left( \VecTrait -  \RootTrait \Multiply \VecOne  \right)}{\NbrSpecies - 1}. \label{eq:estimated-rate-phy}
\end{gather}

For a species $i$ of interest with inter-individual data available, $\RatePopSpi$ can be estimated as the ratio of genetic variance of a trait ($\VarGeneticSpi = \VarPhenotypeSpi \Multiply \HereditySpi$) over genetic diversity of neutrally evolving sequences ($\pi_{\Spi}$).
Averaged across all species, we obtain the mean as:
\begin{gather}
    \EstRatePop = \dfrac{1}{\NbrSpecies}\sum_{i=1}^{\NbrSpecies}\RatePopSpi = \dfrac{1}{\NbrSpecies}\sum_{i=1}^{\NbrSpecies} \frac{  V_{\Trait, i} \Multiply \Heredity_{i}}{ \pi_{i}}. \label{eq:estimated-rate-pop}
\end{gather}

Importantly, the sequence from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow to normalize for diversity driven by mutation rate and population size.

\subsubsection{Bayesian estimate}
We used the software Bayescode

@@

We will assume that the trait genotypic value is encoded by $\NbrLoci$ loci, with each locus contributing to the genotypic value, these effect are Normally distributed with standard deviation $a$ (Fig.~\ref{fig:simulator}).
We will follow a Wright-Fisher model with mutation, selection and drift to formalise the evolutionary dynamics of the system.
Assuming a diploid panmictic population of size $\Ne$ and with non-overlapping generations, we can simulate explicitly each generation along a phylogenetic tree.
Genetic drift is modelled by the multinomial resampling of the currently segregating alleles.
The number of mutations is drawn from a Poisson distribution with mean $2 \Multiply \MutationRate \Multiply \NbrLoci $.
The changes in the mutation rate per generation ($\MutationRate$) and $\Ne$ along the lineages is modelled by a geometric Brownian process.
Additionally, the instant value of log-$\Ne$ is modelled as a sum of a geometric Brownian process and an Ornstein-Uhlenbeck process.
The geometric Brownian motion accounts for long-term fluctuations, while the Ornstein-Uhlenbeck introduces short-term fluctuations.
The simulations will start from an initial sequence at equilibrium at the root of the tree and, at each node, the process will be split until it finally reaches the leaves of the tree.
From a speciation process perspective, this a equivalent to an allopatric speciation over one generation.
Selection is modelled as a one dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and optimal phenotype \cite{tenaillon_utility_2014,blanquart_epistasis_2016}.
The exact functional phenotype-fitness map depends on $2$ external parameters controlling for strength of selection and optimum phenotype.
Mutations are seen has displacement of the phenotype in the multidimensional space.
Beneficial mutations moves the phenotype closer to the optimum, while deleterious mutations moves it further away.
Stabilizing selection is implemented by means of a fixed optimum phenotype.
Diversifying selection is implemented by means of a moving optimum phenotype, changing along the phylogeny as a geometric Brownian process.

@@

\begin{table*}[t!]
    \centering
    \begin{adjustbox}{width = 1.0\textwidth}
    \begin{tabular}{||l|l||c|c||c|c||}
            \hline
            Assumption & Consequences & $\EstRatePop$ & $\EstRatePhy$ & Test $\NIx > 1$  & Test $\NIx < 1$ \\ \hline \hline
            Changing $\MutationRate$ & Scaling with $\pi$, $d$, $\VarPhenotype$ and $\VarPhy$ & Adjusted & Adjusted & -- & --  \\ \hline
            Changing $\Ne$ & Scaling with $\pi$ and $\VarPhenotype$ & Adjusted & -- & -- & --  \\ \hline
            Unknown heredity & $\Heredity = 1$ is over-estimated  & Over-estimated & -- & Conservative & False positives  \\ \hline
            Inbreeding & $\pi$ is under-estimated  & Over-estimated & -- & Conservative & False positives  \\ \hline
            Polymorphism are selected & $\pi$ is under-estimated  & Over-estimated & -- & Conservative & False positives  \\ \hline
            Sexual dimorphism & $\VarPhenotype$ is over-estimated  & Over-estimated & -- & Conservative & False positives  \\ \hline
            Trait encoded by few loci & $\VarPhy$ is under-estimated  & -- & Under-estimated & Conservative & False positives  \\ \hline
            Divergence are selected & $d$ is under-estimated & -- & Over-estimated & False positives & Conservative  \\ \hline
            Structured population & unknown & unknown & unknown & unknown & unknown \\ \hline
            Gene flow between species & unknown & unknown & unknown & unknown & unknown \\ \hline
    \end{tabular}
    \end{adjustbox}
    \caption{Assumptions and their consequences on deviation from neutraliity.}
    \label{table:assumptions}
\end{table*}

Importantly, the genetic sequences from which $\pi$ and $d$ are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow to normalize for diversity driven by mutation rate and population size.
Analogously to the \textcite{mcdonald_adaptative_1991} test devised for protein-coding DNA sequences, a category of neutral mutations (synonymous) is used to determine the neutral expectation, and the inflation of divergence is compared to polymorphism within species.
The relationship between trait variation between and within species can be used as method to detect whether traits are evolving under a neutral model of evolution.
Similarly, in molecular evolution, non-synonymous divergence ($\dn$) and polymorphism $\pn$ are normalized by synonymous divergence ($\ds$) and polymorphism ($\ps$) to produce a neutrality index\cite{mcdonald_adaptative_1991, fay_evaluating_2008}.
Altogether, our neutrality index for a quantitative trait leverages the data for any number of species, and takes advantage of the signal over the whole phylogeny, while at the same time taking into account phylogenetic inertia and addressing the non-independence between species (see fig.~\ref{fig:methods}).
This statistic can be obtained as a maximum likelihood estimate, from eq.~\ref{eq:estimated-rate-pop} and~\ref{eq:estimated-rate-phy}.

@@

In this study, we proposed a neutrality index for a quantitative trait.
At the phylogenetic scale, trait variation between species is normalized by sequence divergence obtained from a neutral set of markers.
Similarly, trait variation within species is normalized by sequence polymorphism obtained from a neutral set of markers.
Our neutrality index for a trait is the ratio of between species variation to within species variation.
Such neutrality index can be compared to 1 to seek a statistical deviation from a neutral trait.
In other words, our statistical test does to seeks to find traits with extremal values in comparison to a pool of traits.
Instead, our procedure can be applied to a single trait, estimating the neutrality index and giving a statistical test for departure from the null model of neutral evolution.
Technically, the neutrality index is estimated either as a maximum likelihood point estimate, or as mean posterior from a Bayesian implementation also giving access to the posterior credible interval allowing to test for a departure from a neutrally evolving trait.
Importantly, our neutrality index leverages sequence divergence and polymorphism, as well as the heritability of traits.
We argue that the main novelty of this study is being able to use sequence divergence and polymorphism to normalize the variation of trait between and within species.
By doing so, our test is not sensitive to the assumption that population sizes ($\Ne$) and mutation rates ($\MutationRate$) are constant across the phylogeny, not requiring to estimate divergence time\cite{litsios_effects_2012}.
Instead, the normalization by sequence divergence and polymorphism automatically absorbs these changes\cite{seo_estimating_2004}.
Contrarily, we argue that the main drawback of our method is that the heritability ($\Heredity$) of a trait is required if one has access only to the phenotypic variance ($\VarPhenotype$) and not the genetic variance ($\VarGenetic$), which is empirically hard to measure for any trait.

% What about the other methods? What are we testing for?
At the phylogenetic scale, the change in mean trait value across different species is typically used to test for neutrality for a quantitative trait.
These models are seeking deviation from a Brownian process\cite{catalan_drift_2019}, or from an Ornstein–Uhlenbeck process accounting for within-species variation\cite{rohlfs_phylogenetic_2015}.
For example, the analysis on the transcriptome across different individuals and species concluded that the expression level of 81\% of genes are under a neutral regime and 16\% under directional selection\cite{catalan_drift_2019}.
However, the alternative model of a varying directional or stabilizing selection was not implemented, and thus can not be ruled out by this analysis.
With our statistical test, we can set a threshold for neutrality and thus are able to make the difference between a trait evolving neutrally as a Brownian process or a trait under selection where the optimal is moving as a Brownian.
Our diversity index can thus allow to revisit these studies, assuming we have access to genomic datasets to estimate divergence ($d$) and polymorphism ($\pi$), as well as estimate for heritabilities ($\Heredity$).

@@

\part*{Supplementary materials}

\section{Theoretical derivation}
\subsection{Definitions}

We denote $\Trait_{i}$ the quantitative trait $\Trait$ for individual $i$.

\begin{align}
    \VarPhenotype = \sum_{i=1}^{\Ne} \Trait_{i}^2 - \left( \sum_{i=1}^{\Ne} \Trait_{i} \right)^2
\end{align}
